---
title: "SHAPmap Visualization Workflow"
author: "Shad Arif Mohammed"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SHAPmap Visualization Workflow}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
 knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>"
 )
```

# Introduction

This vignette demonstrates the complete workflow for generating **SHAP (SHapley Additive exPlanations) value spatial heatmaps** from mass spectrometry imaging (MSI) data. The workflow trains an XGBoost binary classifier, computes SHAP values using the `treeshap` package, and visualizes the results spatially on the MSI raster.

## Prerequisites

The following R packages are required:

- **treeshap**: For computing SHAP values from tree-based models
- **xgboost**: For training the gradient boosting classifier  
- **moleculaR**: For MSI-specific functions (spatial window creation)
- **data.table**: For efficient data manipulation
- **viridisLite**: For color palettes

```{r load-packages, message = FALSE}
# Load required packages
library(treeshap)
library(xgboost)
library(moleculaR)
library(data.table)
library(viridisLite)
```

---

# Step 1: Load Input Data

## Overview

We load the MSI training data from the `trainSet_spmat.rds` file, which contains:
- `$spmat`: Sparse intensity matrix (pixels × m/z features)
- `$coordinates`: Pixel coordinates (x, y positions)
- `$mzAxis`: m/z values for each feature column

## Data Structure

The data is stored as a sparse matrix in dgCMatrix format for memory efficiency. Each row represents a pixel, and each column represents an m/z feature.

```{r load-data}
# Define paths
abs_path <- "D:/Projects/Rpackages/SIFS/"
input_dir <- file.path(abs_path, "data/input")
output_dir <- file.path(abs_path, "data/output/workflow2")

# Load the training dataset
train_spmat <- readRDS(file = file.path(input_dir, "trainSet_spmat.rds"))

# Inspect the structure
cat("=== trainSet_spmat Structure ===\n")
cat("spmat dimensions:", dim(train_spmat$spmat), "\n")
cat("coordinates: ", nrow(train_spmat$coordinates), "pixels\n")
cat("mzAxis: ", length(train_spmat$mzAxis), "m/z features\n")

# Extract components for easier access
spmat <- train_spmat$spmat
coords <- train_spmat$coordinates
mz_axis <- train_spmat$mzAxis

cat("\n=== Coordinate Ranges ===\n")
cat("X range:", min(coords$x), "-", max(coords$x), "\n")
cat("Y range:", min(coords$y), "-", max(coords$y), "\n")
```

**Key Points:**
- The coordinate system uses pixel indices (x: 788-878, y: 138-189)
- Total of 3,660 pixels in the tissue sample
- 13,144 m/z features detected

---

# Step 2: Prepare Training Data

## Overview

Convert the sparse matrix to dense format and prepare feature matrix (X) and labels (y) for model training.

```{r prepare-data}
# Convert sparse matrix to dense matrix
X_train <- as.matrix(spmat)

# Set column names to m/z values (required for treeshap compatibility)
colnames(X_train) <- as.character(mz_axis)

# Load labels (binary class: tumor vs normal)
labels <- read.csv(file = file.path(input_dir, "t3_lbls.csv"))
y_train <- labels$t3_labels

cat("=== Training Data ===\n")
cat("X_train dimensions:", dim(X_train), "\n")
cat("Labels:", length(y_train), "\n")
cat("Class distribution:\n")
print(table(y_train))
```

**Important Note:**
We use the **original m/z values** from `mz_axis`, NOT the rounded column names. The m/z values directly from the data preserve full precision.

---

# Step 3: Train XGBoost Classifier

## Overview

Train an XGBoost gradient boosting classifier for binary classification. We use simplified parameters for faster computation while maintaining reasonable accuracy.

```{r train-model, message = FALSE}
# XGBoost training parameters
params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 6,
  eta = 0.1,
  nthread = 2,
  verbosity = 0
)

# Train the model
cat("Training XGBoost model...\n")
xgb_model <- xgboost(
  data = X_train,
  label = y_train,
  params = params,
  nrounds = 100,
  verbose = FALSE
)

# Check training accuracy
train_pred <- predict(xgb_model, X_train)
Binarized_train_pred <- ifelse(train_pred < 0.5, yes = 0, no = 1)
train_auc <- pROC::auc(Binarized_train_pred, y_train)
cat(sprintf("Training AUC: %.4f\n", train_auc))
```

---

# Step 4: Compute SHAP Values

## Overview

Use the `treeshap` package to compute SHAP values. This requires:

1. **Unify** the model using `xgboost.unify()`
2. Compute SHAP values with `treeshap()`

The unified model format is required by treeshap to compute SHAP values efficiently. The SHAP values are stored in the `$shaps` component of the results.

```{r compute-shap, message = FALSE}
cat("Computing SHAP values...\n")

# Step 4a: Unify the XGBoost model
# This converts the model to treeshap's required format
unified_model <- xgboost.unify(xgb_model, X_train)

# Step 4b: Compute SHAP values
# verbose = FALSE suppresses progress output
shap_results <- treeshap(unified_model, X_train, verbose = FALSE)

# Extract SHAP values from the $shaps component
shap_values <- shap_results$shaps

cat("SHAP values dimensions:", nrow(shap_values), "x", ncol(shap_values), "\n")
cat("SHAP values class:", class(shap_values), "\n")
```

---

# Step 5: Calculate Global SHAP Importance

## Overview

Compute the mean absolute SHAP value for each feature to determine global importance. Rank features by importance to identify the top contributors to classification.

```{r global-importance}
# Set feature names as m/z values
colnames(shap_values) <- as.character(mz_axis)

# Calculate mean absolute SHAP for each feature
mean_abs_shap <- colMeans(abs(shap_values))

# Create importance dataframe
global_shap_df <- data.frame(
  feature = names(mean_abs_shap),
  mean_abs_shap = as.numeric(mean_abs_shap)
)

# Sort by importance (descending)
global_shap_df <- global_shap_df[order(-global_shap_df$mean_abs_shap), ]

# Get top 10 features
top10_mz <- global_shap_df$feature[1:10]

cat("=== Top 10 m/z by Mean |SHAP| ===\n")
print(top10_mz)

# Save to CSV
fwrite(global_shap_df, file = file.path(output_dir, "global_shap_importance.csv"))
```

---

# Step 6: Create Local SHAP Contributions

## Overview

For each pixel, extract the SHAP contribution values for the top 10 features. This creates the local SHAP matrix used for spatial visualization.

```{r local-contributions}
# Select top 10 features
local_shap_df <- as.data.frame(shap_values[, top10_mz])

# Add proper column names
colnames(local_shap_df) <- top10_mz

# Save to CSV
fwrite(local_shap_df, file = file.path(output_dir, "local_shap_contributions.csv"))

cat("Local SHAP dimensions:", dim(local_shap_df), "\n")
cat("Saved to:", file.path(output_dir, "local_shap_contributions.csv"), "\n")
```

---

# Step 7: Generate Spatial SHAP Heatmaps

## Overview

Create spatial heatmap visualizations (SHAPmaps) for each of the top 10 m/z features. This workflow:

1. Creates a matrix matching the pixel coordinate system
2. Maps SHAP values to spatial positions
3. Applies a diverging color palette
4. Overlays the tissue boundary

## Understanding the Coordinate System

In R's `image()` function:
- **Matrix rows** correspond to **y-axis** (vertical direction)
- **Matrix columns** correspond to **x-axis** (horizontal direction)
- `z[row, col]` → `image(x, y, z)` where `length(x) = ncol(z)` and `length(y) = nrow(z)`

## Matrix Construction

The matrix is constructed such that:
- Rows correspond to y-coordinates (52 unique values: 138-189)
- Columns correspond to x-coordinates (91 unique values: 788-878)

```{r generate-heatmaps}
# Create output directory
shapmap_dir <- file.path(output_dir, "SHAPmaps")
dir.create(shapmap_dir, recursive = TRUE, showWarnings = FALSE)

# Get coordinate ranges
min_x <- min(coords$x)
min_y <- min(coords$y)
max_x <- max(coords$x)
max_y <- max(coords$y)

nx <- as.integer(max_x - min_x + 1)  # 91
ny <- as.integer(max_y - min_y + 1)  # 52

cat(sprintf("Raster dimensions: nx=%d, ny=%d\n", nx, ny))

# Create spatial window for boundary
spwin <- createSpatialWindow(pixelCoords = coords)
bdry <- spwin$bdry[[1]]

# Custom color palette: blue -> magenta -> red
pastelPalette <- colorRampPalette(c("#018bfb", "#027af4", "#ae12a5", "#cc0095", "#ff024f"))

# Colorbar helper function
add_colorbar <- function(col, breaks_vals, title) {
  usr <- par("usr")
  xpad <- (usr[2] - usr[1]) * 0.02
  xloc <- usr[2] + xpad
  yloc <- seq(usr[3], usr[4], length.out = length(col) + 1)
  rect(rep(xloc, length(col)), yloc[-length(yloc)],
       rep(xloc + xpad * 2, length(col)), yloc[-1],
       col = col, border = NA)
  axis(side = 4, at = yloc, labels = breaks_vals, las = 1, cex.axis = 0.7)
  mtext(side = 4, title, at = mean(yloc), line = 2.5, las = 0, cex = 0.8)
}

# Generate heatmaps for top 10 features
for (i in 1:10) {
  mz_val <- top10_mz[i]
  
  # Get SHAP values for this feature
  shap_vals <- local_shap_df[[as.character(mz_val)]]
  
  # Calculate quantile bounds for color scaling
  q05 <- quantile(shap_vals, 0.05)
  q95 <- quantile(shap_vals, 0.95)
  shap_range <- max(abs(q05), abs(q95))
  
  # Create matrix: rows = x (91), columns = y (52) to match image() expectations
  # image() expects: nrow(z) = length(y), ncol(z) = length(x)
  mat <- matrix(NA, nrow = nx, ncol = ny)
  
  for (j in 1:nrow(coords)) {
    xi <- coords$x[j] - min_x + 1
    yi <- coords$y[j] - min_y + 1
    if (xi >= 1 && xi <= nx && yi >= 1 && yi <= ny) {
      mat[xi, yi] <- shap_vals[j]
    }
  }
  
  # Define output file
  png_file <- file.path(shapmap_dir, sprintf("SHAPmap_%02d_mz_%.3f.svg", i, as.numeric(mz_val)))
  
  # Create plot
  svg(png_file, width = 12, height = 10)
  par(mar = c(4, 4, 3, 6), oma = c(0, 0, 2, 0))
  
  # Color settings
  ncol <- 100
  cols <- pastelPalette(ncol)
  breaks_vals <- seq(-shap_range, shap_range, length.out = ncol + 1)
  
  # Axis vectors
  x_axis <- seq(min_x, max_x, length.out = nx)
  y_axis <- seq(min_y, max_y, length.out = ny)
  
  # Create heatmap
  image(x = x_axis, y = y_axis, z = mat, col = cols,
        breaks = breaks_vals,
        xlab = "X (pixels)", ylab = "Y (pixels)",
        main = "", useRaster = TRUE)
  
  # Add tissue boundary
  lines(bdry$x, bdry$y, col = "black", lwd = 2.5)
  
# Add title
mtext(sprintf("m/z = %.3f (Rank #%d)", as.numeric(mz_val), i), side = 3, outer = TRUE,
        cex = 1.5, font = 2, line = 0)
  
  # Add colorbar
  add_colorbar(col = cols, breaks_vals = round(breaks_vals, 2), title = "SHAP")
  
  dev.off()
  
  cat(sprintf("Generated: %s\n", basename(png_file)))
}

cat(sprintf("\n=== Complete! %d SHAPmaps saved to %s ===\n", 10, shapmap_dir))
```

---

# Understanding the Output

## SHAPmap Interpretation

Each SHAPmap shows the spatial distribution of SHAP values for a specific m/z feature:

- **Red colors** (right side of palette): Positive SHAP values → feature pushes prediction toward class 1 (tumor)
- **Blue colors** (left side of palette): Negative SHAP values → feature pushes prediction toward class 0 (normal)
- **Intensity**: Magnitude of the SHAP value indicates feature importance at that pixel

## Feature Selection

The top 10 features are ranked by **mean |SHAP|** across all pixels, representing the most important m/z values for classification overall.

## Spatial Patterns

The heatmaps reveal:
- **Localized patterns**: Where specific m/z features are important
- **Tissue heterogeneity**: Spatial variation in feature relevance
- **Boundary effects**: Feature importance near tissue edges

---

# Session Information

```{r session-info}
sessionInfo()
```

---

# References

- Lundberg, S. M., & Lee, S. I. (2017). A unified approach to interpreting model predictions. *Advances in Neural Information Processing Systems*, 30. 
- SHAP paper DOI:  https://doi.org/10.48550/arXiv.1705.07874

- TreeSHAP: https://github.com/ModelOriented/treeshap
- XGBoost: https://xgboost.readthedocs.io/
- moleculaR: https://github.com/CeMOS-Mannheim/moleculaR
- SIFS : https://github.com/CeMOS-Mannheim/SIFS
